---
title: "Calculating CO2 Content of Blood"
author: "B Akenzua-Sanderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Calculating CO2 Content of Blood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)            
library(dplyr)
library(co2ntent)

# define a function for compiling Bland-Altman plots
bland_altman_plot <- function(
  data,
  var_a,
  var_b,
  plot_title="Bland-Altman Plot of var_a vs. var_b",
  xlab = "Mean (var_a + var_b)/2",
  ylab = "Difference in measurements (var_a - var_b)",
  mean_colour ="blue",
  sd_colour = "red"
) {
  # calclulate some data points
  data <- data |>
    mutate(
      mean = ({{ var_a }} + {{ var_b }}) / 2,
      diff = {{ var_a }} - {{ var_b }}
    )
  mean_diff <- mean(data$diff)
  sd_diff <- sd(data$diff)
  min_mean <- min(data$mean)
  max_mean <- max(data$mean)
  ul <- mean_diff + (1.96 * sd_diff)
  ll <- mean_diff - (1.96 * sd_diff)
  x_labels <- min_mean - 0.1 * (max_mean -min_mean)
  x_lim <- c(min_mean - 0.2 * (max_mean - min_mean), max_mean + 0.1 * (max_mean - min_mean))

  # construct plot
  plot <- ggplot(data, aes(mean, diff)) +
    geom_hline(yintercept = mean_diff, colour=mean_colour) +
    annotate("text", x= x_labels, y = mean_diff, label=glue::glue("Mean = {d}", d=round(mean_diff,2)), vjust=1.5, colour=mean_colour) +
    geom_hline(yintercept = ul, colour=sd_colour) +
    annotate("text", x= x_labels, y = ul, label=glue::glue("+1.96 SD = {d}", d=round(ul, 2)), vjust=-0.5, colour=sd_colour) +
    geom_hline(yintercept = ll, colour=sd_colour) +
    annotate("text", x= x_labels, y = ll, label=glue::glue("-1.96 SD = {d}", d=round(ll, 2)), vjust=1.5, colour=sd_colour) +
    geom_point(shape=1, size=4) +
    coord_cartesian(xlim = x_lim) +
    labs(
      x=xlab, y=ylab, title = plot_title
    )
}
```

# Introduction

Several canonical methods of calculating the CO~2~ content in blood are described. `co2ntent` currently implements three methods:

-   @douglas_1988
-   @siggaard_1988
-   @loeppky_1983

Using the data from Table 3 from @douglas_1988, this vignette will demonstrate the calculation of the CO~2~ content of whole blood ($Ct_{CO_2}$)and compare each method to measured values of $Ct_{CO_2}$. Measurements wre performed using the Van Slyke method of determining total CO~2~ content ($Ct_{CO_2.VS}$) by vacuum extraction; releasing gases from a blood sample using acid, followed by manometric measurement.

```{r}

# Calculate CO2 content via the three canonical methods
df <- co2ntent::douglas_table_3 |> 
  mutate(
    douglas_co2_blood_ct_ml_dl = co2_content(
      pco2 = pco2_torr,
      ph = ph,
      haemoglobin_g_dl = haemoglobin_g_dl,
      so2_fraction = so2_fraction,
      phase = "blood",
      method = "douglas",
      content_units = "ml/dL",
      pressure_units = "mmHg"
    ),
    siggaard_co2_blood_ct_ml_dl = co2_content(
      # calculate hco3
      hco3_mmols_l = purrr::map2_dbl(
        pco2_torr,
        ph,
        actual_bicarbonate_content_mmol_l,
        pco2_units = "mmHg"
      ),
      pco2 = pco2_torr,
      haemoglobin_g_dl = haemoglobin_g_dl,
      so2_fraction = so2_fraction,
      ph = ph,
      phase = "blood",
      method = "siggaard_andersen",
      content_units = "ml/dL",
      pressure_units = "mmHg"
    ),
    loeppky_co2_blood_ct_ml_dl = co2_content(
      pco2 = pco2_torr,
      phase = "blood",
      method = "loeppky",
      content_units = "ml/dL",
      pressure_units = "mmHg"
    )
  )
```

# Methods and theoretical background

# Douglas method

The content of CO~2~ in blood calculated via the method developed by @douglas_1988 is termed $Ct_{CO_2.Douglas}$

```{r}
bland_altman_plot(
  data=df, 
  var_a=blood_co2_content_ml_dl, 
  var_b=douglas_co2_blood_ct_ml_dl,
  xlab = "Mean (Ct + Douglas)/2",
  ylab = "Difference (Van Slycke - Douglas)",
  plot_title = ""
  ) +
  theme_minimal()
```

# Siggard-Andersen method

The content of CO~2~ in blood calculated via the method developed by @siggaard_1988 is termed $Ct_{CO_2.Siggard}$

```{r}
bland_altman_plot(
  data=df, 
  var_a=blood_co2_content_ml_dl, 
  var_b=siggaard_co2_blood_ct_ml_dl,
  xlab = "Mean (Ct + Siggard)/2",
  ylab = "Difference (Van Slycke - Siggard)",
  plot_title = ""
  ) +
  theme_minimal()
```

# Loeppky method

The content of CO~2~ in blood calculated via the method developed by @loeppky_1983 is termed $Ct_{CO_2.Loeppky}$

```{r}
bland_altman_plot(
  data=df, 
  var_a=blood_co2_content_ml_dl, 
  var_b=loeppky_co2_blood_ct_ml_dl,
  xlab = "Mean (Ct + Loeppky)/2",
  ylab = "Difference (Van Slycke - Loeppky)",
  plot_title = ""
  ) +
  theme_minimal()
```

## References